---
title: "Reading Profile Analysis"
subtitle: "N = 1,504 (grades 2-5)"
author: "Yukie Toyama"
date: "2026-94-17"
output: 
  html_document:
    toc: yes
editor_options: 
  markdown: 
    wrap: 72
---

##  {.tabset}

### Intro
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,    # hide code
  message = FALSE, # hide messages
  warning = FALSE  # hide warnings
)
```

```{r library-loads}
library(mclust)
library(tidyverse)
library(dplyr)
library(corrplot)

```

We conducted latent profile analysis. Initial dataset includes students in grades 2-8 but we subsetted it to include a total of 1,504 students in **grades 2-5** who completed eight ROAR-measures.

```{r write-out-anonymized-data}
# Load in raw data and write out a simplified and totally de-identified data frame
df <- read_csv("data/deidentified_reading_profile_data_2026-03-03.csv")
varnames <- c("word_theta", "phonics_theta" ,"phoneme_theta", "sentence_score", "morphology_theta", "cva_theta", "trog_theta", "inference_theta", "age_months_at_run", "user_grade_at_run") 
df <- df %>% select(all_of(varnames)) %>%
  round(digits=2)
# write.csv(df, "data/deidentified_reading_profile_data_2026-03-03.csv")
```

```{r data-prep}
# Histogram of grades (K-8)
ggplot(df, aes(x=user_grade_at_run)) +
  geom_histogram() +
  labs(title = "Figure 1. Histogram of students in the original dataset (grades 2-8) who completed all 8 ROAR-measures.")

# create a new matrix with the variables used for clustering and removing any row with an NAN and filtering to grades 2-5
df.mclust <- 
  filter(df, user_grade_at_run>=2) %>%
  filter(user_grade_at_run<=5) %>%
  select(varnames[1:8]) 

df.mclust_na <- filter(df.mclust, 
                       !is.na(rowMeans(df.mclust)))

# scaling data: all ROAR scores on the z-score scale
scaled_scores_na <- df.mclust_na %>% 
  mutate(across(everything(), ~as.numeric(scale(.))))
```

### Correlation


```{r corr_mat, fig.cap="**Figure 2.** *Correlations among eight ROAR measures (N = 1504)*"}
# compute correlation matrix and visualize
c = cor(scaled_scores_na)
corrplot(c, 
         method = 'circle', 
         type = 'lower',    # To show lower half only
         order = 'hclust',  # Keeps the hierarchical clustering order
         addCoef.col = "black", 
         diag = FALSE,          # to hide the diagonal
         number.cex = 0.6, 
         number.digits = 2)
```

#### Observations: 

> **A. Language Block among Inference, Morphology, and CVA:** Inference, morphology, and CVA are forming a tight block with correlations 0.76-0.83, likely forming the latent "language" factor. The strongest relationship is between **CVA** and **Morphology (r = 0.83**). Because this is above 0.80, these variables are statistically "redundant." We expect them to show almost identical patterns in our final profiles.

> **B. The Profile "Sweet Spot"** The correlations between foundational skills (like **Phonics**) and comprehension skills (like **Inference**) sit between *0.50* and *0.65*. This is ideal for Profile Analysis; it means the skills are related, but different enough that we can find students with "uneven" strengths and weaknesses.

> **C. Skill Independence** The lowest correlation is between **Sentence Score** and **TROG** (r=0.46). This suggests these two assessments are measuring different cognitive abilities, which may help us distinguish between different "types" of readers in our profiles.


### Latent Profile Analysis

**On this page:**
- [Step 1: Exploratory Model Selection (mclust)](#step1)
- [Step 2. Model Selection & Fit (tidyLPA)](#step2)

```{r lpa-model-1, echo=FALSE}
# Following these instructions for latent profile analysis https://willhipson.netlify.app/post/latent-profile/latent-profile/ 

# consider tidyLPA instead https://cran.r-project.org/web/packages/tidyLPA/vignettes/Introduction_to_tidyLPA.html

# getting and plotting Bayesian Information Criteria 

BIC <- mclustBIC(scaled_scores_na)
```
#### Step 1. Exploratory Model Selection (mclust){#step1}

We sought to identify optimal student subgroups by conducting a Latent Profile Analysis (LPA) using the Mclust function within the *mclust* package in R. This package utilizes a Gaussian Mixture Modeling framework to identify unobserved (latent) groups within the data. Based on a systematic comparison of 14 different model structures in the **Model Specification Plot (Figure 1)**, we selected a 3-profile solution using a **VVE** (Variable volume, Variable shape, Equal orientation) specification.

This model was chosen because it achieved a peak in the Bayesian Information Criterion (BIC), indicating a superior balance between model fit and complexity. The VVE structure is particularly well-suited for reading data as it allows for subgroups of different sizes and different levels of score-variance, reflecting the reality that struggling reader groups are often smaller and more heterogeneous than typical reader groups. This selection process ensures the identified profiles represent meaningful, distinct patterns of reading skills rather than statistical noise

```{r bic-plot, fig.cap="**Figure 1.** *BIC Model Selection Plot. Higher values indicate better fit. The 3-profile VVE model (purple line) was selected as the optimal balance of fit and parsimony.*"}
plot(BIC)
```

A comparison of fit indices (see below) revealed that the VVE **(Variable volume, Variable shape, Equal orientation)** model structure consistently outperformed other specifications. While the absolute best statistical fit was observed for a 6-profile solution (BIC = -24,599.32), we selected the 3-profile VVE model (BIC = -24,640.68) for the final analysis.

This decision was based on the principle of parsimony; the 3-profile solution represents a significant "elbow" in the model selection plot, capturing the primary developmental leaps in reading skills without the risk of overfitting the data to minor variations. Furthermore, the 3-profile solution provided a more theoretically coherent framework for identifying distinct instructional needs, a priority highlighted in foundational reading research (e.g., Vargas et al., 2024; Kulesz et al., 2024).

```{r bic-summary}
summary(BIC)
```

**Validation via Integrated Completed Likelihood (ICL)**

To further validate the number of profiles, we examined the Integrated Completed Likelihood (ICL). While BIC can sometimes favor more complex models, ICL includes an additional penalty for "classification entropy" (how clearly students are assigned to a single group).

As shown in the ICL Model Selection Plot (Figure 2), fit indices for the top-performing model structures (VEE, VVE, and VEV) all reached a distinct peak at the 3-profile solution (Top ICL = -25,034.6) before declining. This convergence between BIC and ICL strongly supports the 3-profile model as the most reliable representation of student reading patterns, ensuring high classification certainty and minimal overlap between subgroups.

```{r icl}
# double check we chose the best model by looking at Integrated Completed Likelihood 
ICL <- mclustICL(scaled_scores_na)
```

```{r icl_plot, fig.cap="**Figure 3.** *BIC Model Selection Plot. Higher values indicate better fit. The 3-profile VVE model (purple line) was selected as the optimal balance of fit and parsimony.*"}
plot(ICL)
```

```{r icl_sum}
summary(ICL)
```

##### Three Profiles (VVE)

The 3-profile VVE model successfully categorized the 1,504 students into distinct subgroups based on their reading and language profiles. The largest subgroup (Profile 1, n = 791) represented 52.6% of the sample, followed by Profile 3 (n = 595, 39.6%). A smaller, distinct subgroup was also identified (Profile 2, n = 118, 7.8%), representing a specialized pattern of reading development. The high ICL value (-25,061.57) confirms that these three groups are well-separated with high classification certainty.

```{r mod1_3}
#choosing VVE, 3 which is a top model based on BIC
mod1 <- Mclust(scaled_scores_na, modelNames = "VVE", G = 3)
summary(mod1)

# final check Boostrap Likelihood Ratio Test (BLRT) -- takes a long time to run do not run unless necessary 
#mclustBootstrapLRT(scores_na, modelName = "VVV")
```

```{r visualize-lpa, echo=FALSE, warning=FALSE}
# Create a new dataframe that is the mean of each (standardized) measure for each profile
means <- data.frame(mod1$parameters$mean) %>%
  rownames_to_column() %>%
  rename(Scores = rowname) %>%
  pivot_longer(cols = c(X1, X2, X3), names_to = "Profile", values_to = "Mean") %>%
  mutate(Mean = round(Mean, 2)) %>%
  mutate(Scores = factor(Scores, levels = varnames))

# Define the new expert-driven order
order <- c("inference_theta", "morphology_theta", "cva_theta",
           "trog_theta", "sentence_score",  "word_theta", 
           "phonics_theta", "phoneme_theta")

# Update the means dataframe
means <- data.frame(mod1$parameters$mean) %>%
  rownames_to_column() %>%
  rename(Scores = rowname) %>%
  pivot_longer(cols = c(X1, X2, X3), names_to = "Profile", values_to = "Mean") %>%
  mutate(Mean = round(Mean, 2)) %>%
  # --- This is the critical line: re-leveling the factor ---
  mutate(Scores = factor(Scores, levels = order))
```

##### Profile Plot (VVE)

Figure 3 shows the strength and weakness of the three latent profiles on the eight ROAR-measures.

```{r profile_plot, fig.cap="**Figure 4.** *Latent Profiles of Reading and Language Skills. Z-Scores relative to sample average (N = 1504)*"}
library(stringr)

# 1. Define your labels with percentages
labels_with_pct <- c(
  "X1" = "Proficient (52.6%)", 
  "X3" = "Low Average (39.6%)", 
  "X2" = "Struggling (esp. Grammar & Phonics) (7.8%)"
)

# 2. Plotting with strict legend control
vve.c3 <- 
  ggplot(means, aes(x=Scores, y=Mean, group = Profile, color = Profile)) +
  geom_point(size = 3) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", alpha = 0.4) +
  # Clean X-axis labels (removing _theta and _score)
  scale_x_discrete(labels = function(x) str_replace_all(x, "_theta|_score", "")) +
  scale_color_manual(
    values = c("X1" = "#2ca02c", "X3" = "#ff7f0e", "X2" = "#d62728"), # Green, Orange, Red
    breaks = c("X1", "X3", "X2"), # the order: Prof -> Low Ave -> Struggling
    labels = labels_with_pct
  ) +
  
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, face = "bold"),
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    y = "Mean Z-Score (Relative to Average)",
    x = "ROAR Measures"
  )

vve.c3

vve.c3.titles <- vve.c3 + 
  labs(
    title = "VVE: 3-Profile Solution",
    subtitle = "Z-scores relative to sample average (N = 1,504)"
  ) +
  guides(color = guide_legend(ncol = 1))


```

**Profile 1 (Green): "The Proficient Readers" (\~53% of sample)**

The Pattern: A relatively flat line sitting comfortably above the 0.5 standard deviation mark.

Interpretation: These students are "balanced." They show no significant discrepancies between high-level language comprehension and low-level code skills. Their reading system is likely fully integrated.

**Profile 2 (Orange): "The Low-Average" (\~40% of sample)**

The Pattern: Also a relatively flat line, but sitting entirely below the sample average (0). Notice how this line "peaks" slightly at **TROG** and **Phonics**. This suggests that while these students are generally low, they have slightly better receptive grammar and phonics knowledge compared to their broader language comprehension (Inference/Morphology/CVA) and phonemic awareness.

Interpretation: These are students who are "getting by" but are consistently performing below their peers.

**Profile 3 (Red): "The Struggling (esp. Grammar & Phonics) " (\~8% of sample)**

The Pattern: All lower than the other two profiles with distinct valleys with trog and phonics, indicating these students struggle mostly with receptive grammar and decoding. Interesting to note that these students appear to have basic phonemic awareness and language comprehension skills as the red line jump up to trail the orange line for these other skills

Interpretation: This groups has specific double weaknesses. They are not just "globally low" like the low-average group; they have massive, specific "drops" in Phonics and Grammar that are out of proportion with their other skills.


#### Step 2. Model Selection & Fit (tidyLPA){#step2}

Following the exploratory model search in *mclust*, which suggested a 3-profile structure, we conducted a formal validation using the *tidyLPA* package. This transition allows for a more parsimonious comparison of fit indices—specifically Entropy and the Bootstrap Likelihood Ratio Test (BLRT)—ensuring that the final profiles are both statistically robust and clearly differentiated. Additionally, while the *mclust* package was used to explore complex covariance structures (like VVE), this tidyLPA approach utilizes a more parsimonious 'Model 1' specification (equal variances and zero covariances), and 'Model 2' (varying variances), along with VVE. Models 1 and 2 provide a more conservative and highly interpretable framework for selecting and reporting the final student profiles.

##### Model Fit Comparison Results

As shown in **Table 2**, the **VVE 3-profile solution** achieved the absolute best statistical fit according to the Bayesian Information Criterion (**BIC = 24,640.68**). However, this model exhibited relatively lower classification clarity (**Entropy = 0.73**), suggesting significant overlap between the identified groups. In contrast, the **Model 2 (varying variances)** solutions with 4 and 5 classes provided a compelling alternative; while their BIC values were higher than the VVE model, they yielded substantially higher classification certainty (**Entropy = 0.87**).

Given our sample size ($N=1504$) and our diagnostic interest in the 'lower-end' of the reader distribution, we examined Model 2 higher-order solutions. The Model 2 framework allows for the heterogeneity typically found in struggling readers, enabling us to identify smaller, clinically relevant subgroups (representing 3–15% of the sample) which could be masked in more parsimonious yet restrictive models.

```{r mclust_fit_indecies}
library(mclust)
library(dplyr)

# 1. Fit direct mclust VVE model with 3 classes
mod_vve_3 <- Mclust(
  data = as.data.frame(scaled_scores_na),
  G = 3,
  modelNames = "VVE"
)

# 2. Extract core quantities
n_obs <- nrow(scaled_scores_na)
logLik_vve <- mod_vve_3$loglik
df_vve <- mod_vve_3$df

# mclust uses the opposite BIC sign convention from tidyLPA
# mclust: larger is better
# tidyLPA table: smaller is better
BIC_vve <- -mod_vve_3$bic

# Optional: AIC in the usual "smaller is better" form
AIC_vve <- -2 * logLik_vve + 2 * df_vve

# 3. Compute entropy from posterior probabilities
z <- mod_vve_3$z
Entropy_vve <- 1 + sum(z * log(z + 1e-10)) / (n_obs * log(ncol(z)))

# 4. Compute smallest class proportion (to match n_min in your table)
class_props <- prop.table(table(mod_vve_3$classification))
n_min_vve <- min(class_props)

# largest class proportion if you also want n_max
n_max_vve <- max(class_props)

# 5. Create one-row summary matching your tidyLPA sensitivity table
vve_fit_row <- tibble(
  Model_Type = "VVE (Variable Shape/Volume)",
  Classes = 3,
  n_parm = df_vve,
  BIC = round(BIC_vve, 2),
  Entropy = round(Entropy_vve, 2),
  n_min = round(n_min_vve, 2),
  BLRT_p = NA_real_
)

# vve_fit_row

```

```{r tidyLPA_mod_com2}

library(tidyLPA)
library(dplyr)
library(kableExtra)

# Note: In mclust terms: 
# Model 1 = EEI (Equal variance, diagonal)
# Model 2 = VVI (Varying variance, diagonal)

# --- Run Model 1 (Equal Variance) ---
mod1_results <- scaled_scores_na %>%
  estimate_profiles(1:5, package = "mclust", 
                    variances = "equal", 
                    covariances = "zero") %>%
  get_fit() %>%
  mutate(Model_Type = "Model 1 (Equal Var)")

# --- Run Model 2 (Varying Variance) ---
mod2_results <- scaled_scores_na %>%
  estimate_profiles(1:5, package = "mclust", 
                    variances = "varying", 
                    covariances = "zero") %>%
  get_fit() %>%
  mutate(Model_Type = "Model 2 (Varying Var)")


# --- Combine into one Master Comparison Table ---
all_comparisons <- bind_rows(mod1_results, mod2_results)

all_comparisons_short <- all_comparisons %>%
  select(Model_Type, Classes, n_parm = parameters,
         BIC, Entropy, n_min, BLRT_p)

all_comparisons_short <- bind_rows(
  all_comparisons_short,
  vve_fit_row
)

# --- Clean and Present ---
all_comparisons_short %>%
  mutate(across(c(BIC, Entropy, n_min), ~round(., 2))) %>%
  arrange(Model_Type, Classes) %>% 
  kbl(caption = "Table 1. Sensitivity Analysis: Comparing Profile Solutions across Model Specifications (Z-Scores)") %>%
  kable_classic(full_width = F, html_font = "Arial") 
# %>%
#   # Highlight all 3-profile solutions to compare the 'Winner' across specifications
#   row_spec(which(all_comparisons$Classes == 3), bold = T, background = "#f9f9f9")

```

##### Four profiles from Model 2 (varying var)

Figure below shows four profiles from Model 2 (varying variances, zero covariances). The colored lines are mostly parallel.

**Highlight**: This solution primarily identifies levels of risk. We see a **proficient** group (green), an **above-average** group (blue), a **mild-risk** group (orange), and a chronic-risk group (red).

**The Diagnostic Takeaway**: The 4-profile solution (Model 2) primarily reflects a unidimensional severity gradient. Because the profiles are largely parallel across all eight indicators, the differences between subgroups appear to be quantitative (differences in degree of skills) rather than qualitative (differences in the pattern of skills). This suggests that within this 4-class framework, the latent space is defined by a singular 'general reading ability' factor where students are distinguished by their overall developmental level.

```{r profile_plots_m2.c4, fig.cap="**Figure 5.** *Model 2: 4-Profile Solution (varying variances). Z-Scores relative to sample average (N = 1504)*"}

# ---Get the Model Results ---
mod2_fitted <- scaled_scores_na %>%
  estimate_profiles(1:5, package = "mclust", 
                    variances = "varying", covariances = "zero")

model2_4class <- mod2_fitted[[4]]

# ---Get Estimates and Calculate % ---
plot_data_4 <- get_estimates(model2_4class) %>%
  dplyr::filter(Category == "Means") %>%
  mutate(Parameter = factor(Parameter, levels = order))

# Calculate Percentages for the legend
counts_4 <- table(get_data(model2_4class)$Class)
pcts_4 <- round(prop.table(counts_4) * 100, 1)

# --- Create the Plot ---
ggplot(plot_data_4, aes(x = Parameter, y = Estimate, group = factor(Class), color = factor(Class))) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_errorbar(aes(ymin = Estimate - 1.96*se, ymax = Estimate + 1.96*se), width = 0.1, alpha = 0.3) +
  scale_x_discrete(labels = function(x) str_replace_all(x, "_theta|_score", "")) +
  scale_color_manual(
    values = c("1"="#2ca02c", "2"="#d62728", "3"="#1f77b4", "4"="#ff7f0e"),
    labels = paste0("Class ", names(pcts_4), " (", pcts_4, "%)")
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, face = "bold"),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(y = "Mean Z-Score (relative to average)",
       x = "ROAR Measures")

```


##### Five profiles from Model 2 (varying var)

Figure below displays the five profiles from Model 2 (varying variances, zero covariances). Unlike the 4-profile solution, the colored lines here exhibit significant **non-parallelism**, particularly within the lower-performing subgroups.

**Highlights**: The plots reveal both levels of severity and **qualitative differences in skill configurations**. In addition to the proficient (green), above-average (blue), and low-average (purple) groups, the model identifies two distinct types of struggling readers: a **specific syntactic-foundational risk** group (orange, 2.9%) and a **global reading risk** group (Red, 13.4%).

**The Diagnostic Takeaway:** The 5-profile solution reveals a more complex **structural pattern** than the 4-profile model. While the higher-performing groups continue to follow a "level-based" trend, the emergence of the orange profile represents a **qualitative shift** in the data. This subgroup is characterized by a **discrepant pattern**: these students demonstrate relatively average performance on the sentence tasks but show severe, disproportionate "valleys" in receptive grammar (TROG) and phonological decoding (phonics).

This suggests that reading risk in this population is not purely a matter of being "behind" in every skill. By moving to a 5-class framework, we transition from a model of **global developmental delay** to one that detects a small group of students with **specific needs**. This granularity is critical for instructional utility, as it distinguishes students who require a general increase in support from those who require specialized interventions targeting specific linguistic and phonological hurdles.

```{r profile_plots_m2.c5, fig.cap="**Figure 6.** *Model 2: 5-Profile Solution (varying variances). Z-Scores relative to sample average (N = 1504)*"}

# --- Extract the 5rd model in your mod2_fitted list ---
model2_5class <- mod2_fitted[[5]] 

# --- Prepare the Data ---
plot_data_5 <- get_estimates(model2_5class) %>%
  dplyr::filter(Category == "Means") %>%
  mutate(Parameter = factor(Parameter, levels = order))

# Calculate Percentages for the legend
counts_5 <- table(get_data(model2_5class)$Class)
pcts_5 <- round(prop.table(counts_5) * 100, 1)

# Create the Plot

m2c5.plot <- ggplot(plot_data_5, 
       aes(x = Parameter, y = Estimate, 
           group = factor(Class), color = factor(Class))) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_errorbar(aes(ymin = Estimate - 1.96*se, ymax = Estimate + 1.96*se), 
                width = 0.1, alpha = 0.3) +
  scale_x_discrete(labels = function(x) str_replace_all(x, "_theta|_score", "")) +
  scale_color_manual(
    # FIXED: Removed the trailing comma after 'orchid4'
    values = c("1"="#2ca02c", "2"="#d62728", "3"="#ff7f0e", "4"="#1f77b4", "5"="orchid4"),
    labels = paste0("Class ", names(pcts_5), " (", pcts_5, "%)")
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, face = "bold"),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(y = "Mean Z-Score (Relative to Average)",
       x = "ROAR Measures")

m2c5.plot

```

### Cross-Model Convergence

Figure 7 below presents the VVE 3-profile solution alongside the Model 2 5-profile solution. Notably, a similar configuration of dual weaknesses emerges in both models, which is characterized by lower performance in receptive grammar (TROG) and phonological decoding (Phonics) relative to sentence-level skills.

In the VVE solution, this pattern appears within the lowest-performing profile (red line), whereas in the 5-profile Model 2 solution it is isolated as a more distinct subgroup (orange). The recurrence of this feature across modeling frameworks (mclust vs. tidyLPA) and covariance specifications (VVE vs. diagonal Model 2) suggests that this contrast reflects a stable aspect of the observed score structure rather than a model-specific artifact.

At the same time, differences in class separation and profile granularity indicate that this pattern is embedded within a broader severity dimension. Thus, the 5-profile solution can be understood as providing a more fine-grained partitioning of variation that is already present, but less differentiated, in the more parsimonious 3-profile model.

```{r figure-comparison, , fig.width=15, fig.height=7, fig.cap="**Figure 7.** *VVE 3-classe solution vs. Model 2 (VVI) 3-class solution*"}

library(ggplot2)
library(patchwork)
library(stringr)

# 1. Create a shared Y-axis limit so the gridlines match exactly
# Adjust these numbers based on your data's actual min/max
y_limits <- c(-2.5, 1.5) 

# 2. Refine the Model 2 Plot
m2_fixed <- ggplot(plot_data_5, aes(x = Parameter, y = Estimate, group = factor(Class), color = factor(Class))) +
  geom_point(size = 2.5) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", alpha = 0.4) +
  geom_errorbar(aes(ymin = Estimate - 1.96*se, ymax = Estimate + 1.96*se), width = 0.1, alpha = 0.3) +
  # Sync the Y-axis across both plots
  coord_cartesian(ylim = y_limits) +
  scale_x_discrete(labels = function(x) str_replace_all(x, "_theta|_score", "")) +
  scale_color_manual(
    values = c("1"="#2ca02c", "2"="#d62728", "3"="#ff7f0e", "4"="#1f77b4", "5"="#9467bd"),
    labels = paste0("Class ", names(pcts_5), " (", pcts_5, "%)")
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1), 
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank() 
  ) +
  labs(
    title = "Model 2 (VVI): 5-Profile Solution",
    x = "ROAR Measures"
  ) +
  guides(color = guide_legend(ncol = 2, override.aes = list(size = 3)))

# 3. Apply similar limits/cleanup to the first plot (vve.c3.titles)
vve_clean <- vve.c3.titles + 
  coord_cartesian(ylim = y_limits) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(y = "Mean Z-Score")

# 4. Combine with Patchwork
combined_final <- (vve_clean | m2_fixed) + 
  plot_layout(widths = c(1, 1)) & 
  theme(
    plot.margin = margin(10, 10, 10, 10),
    legend.box.margin = margin(t = -10) # Pulls legend closer to plot
  )

combined_final

```

#### Cross-Model Classification Analysis

To evaluate the extent to which the VVE 3-profile solution and the Model 2 5-profile solution recover similar latent groupings, we conducted a cross-model classification analysis based on individual-level class assignments. Overall agreement between the two solutions was modest (Adjusted Rand Index = .32), indicating that the models do not identify identical partitions of the sample.

Particular attention was given to the apparent similarity between VVE Class 3 and Model 2 Class 3 with the double weaknesses in the profile plots. Despite their visual resemblance, cross-model classification analyses indicated that these classes do not represent the same subgroup of individuals.

Specifically, Model 2 Class 3 was composed primarily of individuals from the middle-performing VVE class (72.7%), with a smaller contribution from the lowest-performing VVE class (27.3%), and no representation from the highest-performing class. Conversely, only a negligible proportion of individuals in VVE Class 3 (approximately 2%) were assigned to Model 2 Class 3; instead, the majority were distributed across other Model 2 classes, particularly Class 5 (52.4%).

These results indicate that Model 2 Class 3 does not correspond to a direct refinement of the lowest-performing VVE group. Rather, it reflects a recombination of individuals drawn primarily from the middle-performing class, augmented by a subset of lower-performing students. The similarity in profile shape across models therefore appears to arise from a shared contrast in the data—rather than from a stable subgroup with consistent membership across model specifications.


```{r classfication_overlap}
# Cross-model classification overlap:

# 1. mclust VVE (3 classes)  vs  tidyLPA Model 2 (5 classes)
vve_class <- mod1$classification

# inspect class sizes
vve_sizes <- prop.table(table(vve_class))

# 2. Extract class assignments from tidyLPA mod2.5 results
# model2_5class <- mod2_fitted[[5]]
data_m2_5 <- get_data(model2_5class)

# Look at names to confirm the class-assignment column
# names(data_m2_5)
m2_class <- data_m2_5$Class

# inspect class sizes
m2_sizes <- prop.table(table(m2_class))

# 3. Basic checks
# length(vve_class)
# length(m2_class)


# 4. Cross-tabulation of class assignments
overlap_tab <- table(VVE = vve_class, Model2_5 = m2_class)

# Raw counts as a data frame
overlap_counts_df <- as.data.frame(overlap_tab)


# 5. Row-wise and column-wise proportions
# Row-wise: within each VVE class, where do cases go in Model 2?
overlap_row_prop <- prop.table(overlap_tab, margin = 1)
# round(overlap_row_prop, 3)

# Column-wise: within each Model 2 class, where did cases come from in VVE?
overlap_col_prop <- prop.table(overlap_tab, margin = 2)
# round(overlap_col_prop, 3)

# Data-frame versions for easier inspection/export
overlap_row_df <- as.data.frame(round(overlap_row_prop, 3))
overlap_col_df <- as.data.frame(round(overlap_col_prop, 3))
# overlap_row_df
# overlap_col_df


# 6. Adjusted Rand Index (overall agreement)
ari <- adjustedRandIndex(vve_class, m2_class)

# Interpretation guide:
# < .20   = weak agreement
# .20-.50 = modest
# .50-.80 = moderate to strong
# > .80   = very strong


# 7. Identify the dominant VVE source for each Model 2 class
# For each Model 2 class, find which VVE class contributes the largest proportion
dominant_vve_for_m2 <- apply(overlap_col_prop, 2, function(x) {
  c(
    dominant_VVE_class = which.max(x),
    dominant_prop = max(x)
  )
})

dominant_vve_for_m2 <- as.data.frame(t(dominant_vve_for_m2))
names(dominant_vve_for_m2) <- c("dominant_VVE_class", "dominant_prop")

dominant_vve_for_m2$Model2_5_class <- colnames(overlap_col_prop)

dominant_vve_for_m2 <- dominant_vve_for_m2 %>%
  mutate(
    dominant_VVE_class = as.integer(dominant_VVE_class),
    dominant_prop = round(as.numeric(dominant_prop), 3)
  ) %>%
  select(Model2_5_class, dominant_VVE_class, dominant_prop)


# 8. Identify the dominant Model 2 destination for each VVE class

dominant_m2_for_vve <- apply(overlap_row_prop, 1, function(x) {
  c(
    dominant_M2_class = which.max(x),
    dominant_prop = max(x)
  )
})

dominant_m2_for_vve <- as.data.frame(t(dominant_m2_for_vve))
names(dominant_m2_for_vve) <- c("dominant_M2_class", "dominant_prop")

dominant_m2_for_vve$VVE_class <- rownames(overlap_row_prop)

dominant_m2_for_vve <- dominant_m2_for_vve %>%
  mutate(
    dominant_M2_class = as.integer(dominant_M2_class),
    dominant_prop = round(as.numeric(dominant_prop), 3)
  ) %>%
  select(VVE_class, dominant_M2_class, dominant_prop)


# 9. a compact summary object for reporting

overlap_summary <- list(
  ari = ari,
  counts = overlap_tab,
  row_prop = round(overlap_row_prop, 3),
  col_prop = round(overlap_col_prop, 3),
  dominant_vve_for_m2 = dominant_vve_for_m2,
  dominant_m2_for_vve = dominant_m2_for_vve
)

overlap_summary

# TODO
# 1) Examine profiles separately by grade
# 2) Dimensionality analysis

```



