---
title: "Reading Profile Analysis"
subtitle: "N = 1,504 (grades 2-5)"
author: "Yukie Toyama"
date: "2026-94-17"
output: 
  html_document:
    toc: yes
editor_options: 
  markdown: 
    wrap: 72
---

##  {.tabset}

### Intro

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,    # hide code
  message = FALSE, # hide messages
  warning = FALSE  # hide warnings
)
```

```{r library-loads}
library(mclust)
library(tidyverse)
library(dplyr)
library(corrplot)

```

We conducted latent profile analysis. Initial dataset includes students
in grades 2-8 but we subsetted it to include a total of 1,504 students
in **grades 2-5** who completed eight ROAR-measures.

```{r write-out-anonymized-data}
# Load in raw data and write out a simplified and totally de-identified data frame
df <- read_csv("data/deidentified_reading_profile_data_2026-03-03.csv")
varnames <- c("word_theta", "phonics_theta" ,"phoneme_theta", "sentence_score", "morphology_theta", "cva_theta", "trog_theta", "inference_theta", "age_months_at_run", "user_grade_at_run") 
df <- df %>% select(all_of(varnames)) %>%
  round(digits=2)
# write.csv(df, "data/deidentified_reading_profile_data_2026-03-03.csv")

# library(summarytools)
# view(dfSummary(df))
# view(dfSummary(df.mclust))
```

```{r data-prep}
# Histogram of grades (K-8)
ggplot(df, aes(x=user_grade_at_run)) +
  geom_histogram() +
  labs(title = "Figure 1. Histogram of students in the original dataset (grades 2-8) who completed all 8 ROAR-measures.")

# table by grade
library(janitor)
df %>% mutate(grade = factor(user_grade_at_run)) %>% 
  tabyl(grade) %>% adorn_totals() %>% 
  adorn_pct_formatting()

# create a new matrix with the variables used for clustering and removing any row with an NAN and filtering to grades 2-5
df.mclust <- 
  filter(df, user_grade_at_run>=2) %>%
  filter(user_grade_at_run<=5) %>%
  select(varnames[1:8]) 

df.mclust_na <- filter(df.mclust, 
                       !is.na(rowMeans(df.mclust)))
library(janitor)
df %>% filter(user_grade_at_run < 6) %>% 
  tabyl(user_grade_at_run) %>% adorn_totals()
# scaling data: all ROAR scores on the z-score scale
scaled_scores_na <- df.mclust_na %>% 
  mutate(across(everything(), ~as.numeric(scale(.))))
```

### Correlation

Figure 2 shows a correlation matrix among the eight ROAR measures.

```{r corr_mat, fig.cap="**Figure 2.** *Correlations among eight ROAR measures (N = 1504)*"}
# compute correlation matrix and visualize
c = cor(scaled_scores_na)
corrplot(c, 
         method = 'circle', 
         type = 'lower',    # To show lower half only
         order = 'hclust',  # Keeps the hierarchical clustering order
         addCoef.col = "black", 
         diag = FALSE,          # to hide the diagonal
         number.cex = 0.6, 
         number.digits = 2)
```

#### Observations:

> **A. Language Block among Inference, Morphology, and CVA:** Inference,
> morphology, and CVA are forming a tight block with correlations
> 0.76-0.83, likely forming the latent "language" factor. The strongest
> relationship is between **CVA** and **Morphology (r = 0.83**). Because
> this is above 0.80, these variables are statistically "redundant." We
> expect them to show almost identical patterns in our final profiles.

> **B. The Profile "Sweet Spot"** The correlations between foundational
> skills (like **Phonics**) and comprehension skills (like
> **Inference**) sit between *0.50* and *0.65*. This is ideal for
> Profile Analysis; it means the skills are related, but different
> enough that we can find students with "uneven" strengths and
> weaknesses.

> **C. Skill Independence** The lowest correlation is between **Sentence
> Score** and **TROG** (r=0.46). This suggests these two assessments are
> measuring different cognitive abilities, which may help us distinguish
> between different "types" of readers in our profiles.

### Latent Profile Analysis

**On this page:** - [Step 1: Exploratory Model Selection
(mclust)](#step1) - [Step 2. Model Selection & Fit (tidyLPA)](#step2)
<br><br>

```{r lpa-model-1, echo=FALSE}
# Following these instructions for latent profile analysis https://willhipson.netlify.app/post/latent-profile/latent-profile/ 

# consider tidyLPA instead https://cran.r-project.org/web/packages/tidyLPA/vignettes/Introduction_to_tidyLPA.html

# getting and plotting Bayesian Information Criteria 

BIC <- mclustBIC(scaled_scores_na)
```

#### Step 1. Exploratory Model Selection (mclust) {#step1}

We sought to identify optimal student subgroups by conducting a Latent
Profile Analysis (LPA) using the Mclust function within the *mclust*
package in R. This package utilizes a Gaussian Mixture Modeling
framework to identify unobserved (latent) groups within the data. Based
on a systematic comparison of 14 different model structures in the
**Model Specification Plot (Figure 1)**, we selected a 3-profile
solution using a **VVE** (Variable volume, Variable shape, Equal
orientation) specification.

This model was chosen because it achieved a peak in the Bayesian
Information Criterion (BIC), indicating a superior balance between model
fit and complexity. The VVE structure is particularly well-suited for
reading data as it allows for subgroups of different sizes and different
levels of score-variance, reflecting the reality that struggling reader
groups are often smaller and more heterogeneous than typical reader
groups. This selection process ensures the identified profiles represent
meaningful, distinct patterns of reading skills rather than statistical
noise

```{r bic-plot, fig.cap="**Figure 1.** *BIC Model Selection Plot. Higher values indicate better fit. The 3-profile VVE model (purple line) was selected as the optimal balance of fit and parsimony.*"}
plot(BIC)
```

A comparison of fit indices (see below) revealed that the VVE
**(Variable volume, Variable shape, Equal orientation)** model structure
consistently outperformed other specifications. While the absolute best
statistical fit was observed for a 6-profile solution (BIC =
-24,599.32), we selected the 3-profile VVE model (BIC = -24,640.68) for
the final analysis.

This decision was based on the principle of parsimony; the 3-profile
solution represents a significant "elbow" in the model selection plot,
capturing the primary developmental leaps in reading skills without the
risk of overfitting the data to minor variations. Furthermore, the
3-profile solution provided a more theoretically coherent framework for
identifying distinct instructional needs, a priority highlighted in
foundational reading research (e.g., Vargas et al., 2024; Kulesz et al.,
2024).

```{r bic-summary}
summary(BIC)
```

**Validation via Integrated Completed Likelihood (ICL)**

To further validate the number of profiles, we examined the Integrated
Completed Likelihood (ICL). While BIC can sometimes favor more complex
models, ICL includes an additional penalty for "classification entropy"
(how clearly students are assigned to a single group).

As shown in the ICL Model Selection Plot (Figure 2), fit indices for the
top-performing model structures (VEE, VVE, and VEV) all reached a
distinct peak at the 3-profile solution (Top ICL = -25,034.6) before
declining. This convergence between BIC and ICL strongly supports the
3-profile model as the most reliable representation of student reading
patterns, ensuring high classification certainty and minimal overlap
between subgroups.

```{r icl}
# double check we chose the best model by looking at Integrated Completed Likelihood 
ICL <- mclustICL(scaled_scores_na)
```

```{r icl_plot, fig.cap="**Figure 3.** *BIC Model Selection Plot. Higher values indicate better fit. The 3-profile VVE model (purple line) was selected as the optimal balance of fit and parsimony.*"}
plot(ICL)
```

```{r icl_sum}
summary(ICL)
```

##### Three Profiles (VVE)

The 3-profile VVE model successfully categorized the 1,504 students into
distinct subgroups based on their reading and language profiles. The
largest subgroup (Profile 1, n = 791) represented 52.6% of the sample,
followed by Profile 3 (n = 595, 39.6%). A smaller, distinct subgroup was
also identified (Profile 2, n = 118, 7.8%), representing a specialized
pattern of reading development. The high ICL value (-25,061.57) confirms
that these three groups are well-separated with high classification
certainty.

```{r mod1_3}
#choosing VVE, 3 which is a top model based on BIC
mod1 <- Mclust(scaled_scores_na, modelNames = "VVE", G = 3)
summary(mod1)

# final check Boostrap Likelihood Ratio Test (BLRT) -- takes a long time to run do not run unless necessary 
#mclustBootstrapLRT(scores_na, modelName = "VVV")
```

```{r visualize-lpa, echo=FALSE, warning=FALSE}
# Create a new dataframe that is the mean of each (standardized) measure for each profile
means <- data.frame(mod1$parameters$mean) %>%
  rownames_to_column() %>%
  rename(Scores = rowname) %>%
  pivot_longer(cols = c(X1, X2, X3), names_to = "Profile", values_to = "Mean") %>%
  mutate(Mean = round(Mean, 2)) %>%
  mutate(Scores = factor(Scores, levels = varnames))

# Define the new expert-driven order
order <- c("inference_theta", "morphology_theta", "cva_theta",
           "trog_theta", "sentence_score",  "word_theta", 
           "phonics_theta", "phoneme_theta")

# Update the means dataframe
means <- data.frame(mod1$parameters$mean) %>%
  rownames_to_column() %>%
  rename(Scores = rowname) %>%
  pivot_longer(cols = c(X1, X2, X3), names_to = "Profile", values_to = "Mean") %>%
  mutate(Mean = round(Mean, 2)) %>%
  # --- This is the critical line: re-leveling the factor ---
  mutate(Scores = factor(Scores, levels = order))
```

##### Profile Plot (VVE)

Figure 3 shows the strength and weakness of the three latent profiles on
the eight ROAR-measures.

```{r profile_plot, fig.width=10, fig.height=5, fig.cap="**Figure 4.** *Latent Profiles of Reading and Language Skills. Z-Scores relative to sample average (N = 1504)*"}
library(stringr)

# 1. Define your labels with percentages
labels_with_pct <- c(
  "X1" = "Class 1 (52.6%)", 
  "X3" = "Class 3 (39.6%)", 
  "X2" = "Class 2 (esp. Grammar & Phonics) (7.8%)"
)

# 2. Plotting with strict legend control
vve.c3 <- 
  ggplot(means, aes(x=Scores, y=Mean, group = Profile, color = Profile)) +
  geom_point(size = 3) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", alpha = 0.4) +
  # Clean X-axis labels (removing _theta and _score)
  scale_x_discrete(labels = function(x) str_replace_all(x, "_theta|_score", "")) +
  scale_color_manual(
    values = c("X1" = "#2ca02c", "X3" = "#ff7f0e", "X2" = "#d62728"), # Green, Orange, Red
    breaks = c("X1", "X3", "X2"), # the order: Prof -> Low Ave -> Struggling
    labels = labels_with_pct
  ) +
  
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, face = "bold"),
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    y = "Mean Z-Score (Relative to Average)",
    x = "ROAR Measures"
  )

vve.c3

vve.c3.titles <- vve.c3 + 
  labs(
    title = "VVE: 3-Profile Solution",
    subtitle = "Z-scores relative to sample average (N = 1,504)"
  ) +
  guides(color = guide_legend(ncol = 1))


```

**Profile 1 (Green): "The Proficient Readers" (\~53% of sample)**

The Pattern: A relatively flat line sitting comfortably above the 0.5
standard deviation mark.

Interpretation: These students are "balanced." They show no significant
discrepancies between high-level language comprehension and low-level
code skills. Their reading system is likely fully integrated.

**Profile 2 (Orange): "The Low-Average" (\~40% of sample)**

The Pattern: Also a relatively flat line, but sitting entirely below the
sample average (0). Notice how this line "peaks" slightly at **TROG**
and **Phonics**. This suggests that while these students are generally
low, they have slightly better receptive grammar and phonics knowledge
compared to their broader language comprehension
(Inference/Morphology/CVA) and phonemic awareness.

Interpretation: These are students who are "getting by" but are
consistently performing below their peers.

**Profile 3 (Red): "The Struggling (esp. Grammar & Phonics) " (\~8% of
sample)**

The Pattern: All lower than the other two profiles with distinct valleys
with trog and phonics, indicating these students struggle mostly with
receptive grammar and decoding. Interesting to note that these students
appear to have basic phonemic awareness and language comprehension
skills as the red line jump up to trail the orange line for these other
skills

Interpretation: This groups has specific double weaknesses. They are not
just "globally low" like the low-average group; they have massive,
specific "drops" in Phonics and Grammar that are out of proportion with
their other skills.

#### Step 2. Model Selection & Fit (tidyLPA) {#step2}

Following the exploratory model search in *mclust*, which suggested a
3-profile structure, we conducted a formal validation using the
*tidyLPA* package. This transition allows for a more parsimonious
comparison of fit indices—specifically Entropy and the Bootstrap
Likelihood Ratio Test (BLRT)—ensuring that the final profiles are both
statistically robust and clearly differentiated. Additionally, while the
*mclust* package was used to explore complex covariance structures (like
VVE), this tidyLPA approach utilizes a more parsimonious 'Model 1'
specification (equal variances and zero covariances), and 'Model 2'
(varying variances), along with VVE. Models 1 and 2 provide a more
conservative and highly interpretable framework for selecting and
reporting the final student profiles.

##### Model Fit Comparison Results

As shown in **Table 2**, the **VVE 3-profile solution** achieved the
absolute best statistical fit according to the Bayesian Information
Criterion (**BIC = 24,640.68**). However, this model exhibited
relatively lower classification clarity (**Entropy = 0.73**), suggesting
significant overlap between the identified groups. In contrast, the
**Model 2 (varying variances)** solutions with 4 and 5 classes provided
a compelling alternative; while their BIC values were higher than the
VVE model, they yielded substantially higher classification certainty
(**Entropy = 0.87**).

Given our sample size ($N=1504$) and our diagnostic interest in the
'lower-end' of the reader distribution, we examined Model 2 higher-order
solutions. The Model 2 framework allows for the heterogeneity typically
found in struggling readers, enabling us to identify smaller, clinically
relevant subgroups (representing 3–15% of the sample) which could be
masked in more parsimonious yet restrictive models.

```{r mclust_fit_indecies}
library(mclust)
library(dplyr)

# 1. Fit direct mclust VVE model with 3 classes
mod_vve_3 <- Mclust(
  data = as.data.frame(scaled_scores_na),
  G = 3,
  modelNames = "VVE"
)

# 2. Extract core quantities
n_obs <- nrow(scaled_scores_na)
logLik_vve <- mod_vve_3$loglik
df_vve <- mod_vve_3$df

# mclust uses the opposite BIC sign convention from tidyLPA
# mclust: larger is better
# tidyLPA table: smaller is better
BIC_vve <- -mod_vve_3$bic

# Optional: AIC in the usual "smaller is better" form
AIC_vve <- -2 * logLik_vve + 2 * df_vve

# 3. Compute entropy from posterior probabilities
z <- mod_vve_3$z
Entropy_vve <- 1 + sum(z * log(z + 1e-10)) / (n_obs * log(ncol(z)))

# 4. Compute smallest class proportion (to match n_min in your table)
class_props <- prop.table(table(mod_vve_3$classification))
n_min_vve <- min(class_props)

# largest class proportion if you also want n_max
n_max_vve <- max(class_props)

# 5. Create one-row summary matching your tidyLPA sensitivity table
vve_fit_row <- tibble(
  Model_Type = "VVE (Variable Shape/Volume)",
  Classes = 3,
  n_parm = df_vve,
  BIC = round(BIC_vve, 2),
  Entropy = round(Entropy_vve, 2),
  n_min = round(n_min_vve, 2),
  BLRT_p = NA_real_
)

# vve_fit_row

```

```{r tidyLPA_mod_com2}

library(tidyLPA)
library(dplyr)
library(kableExtra)

# Note: In mclust terms: 
# Model 1 = EEI (Equal variance, diagonal)
# Model 2 = VVI (Varying variance, diagonal)

# --- Run Model 1 (Equal Variance) ---
mod1_results <- scaled_scores_na %>%
  estimate_profiles(1:5, package = "mclust", 
                    variances = "equal", 
                    covariances = "zero") %>%
  get_fit() %>%
  mutate(Model_Type = "Model 1 (Equal Var)")

# --- Run Model 2 (Varying Variance) ---
mod2_results <- scaled_scores_na %>%
  estimate_profiles(1:5, package = "mclust", 
                    variances = "varying", 
                    covariances = "zero") %>%
  get_fit() %>%
  mutate(Model_Type = "Model 2 (Varying Var)")


# --- Combine into one Master Comparison Table ---
all_comparisons <- bind_rows(mod1_results, mod2_results)

all_comparisons_short <- all_comparisons %>%
  select(Model_Type, Classes, n_parm = parameters,
         BIC, Entropy, n_min, BLRT_p)

all_comparisons_short <- bind_rows(
  all_comparisons_short,
  vve_fit_row
)

# --- Clean and Present ---
all_comparisons_short %>%
  mutate(across(c(BIC, Entropy, n_min), ~round(., 2))) %>%
  arrange(Model_Type, Classes) %>% 
  kbl(caption = "Table 1. Sensitivity Analysis: Comparing Profile Solutions across Model Specifications (Z-Scores)") %>%
  kable_classic(full_width = F, html_font = "Arial") 
# %>%
#   # Highlight all 3-profile solutions to compare the 'Winner' across specifications
#   row_spec(which(all_comparisons$Classes == 3), bold = T, background = "#f9f9f9")

```

##### Four profiles from Model 2 (varying var)

Figure below shows four profiles from Model 2 (varying variances, zero
covariances). The colored lines are mostly parallel.

**Highlight**: This solution primarily identifies levels of risk. We see
a **proficient** group (green), an **above-average** group (blue), a
**mild-risk** group (orange), and a chronic-risk group (red).

**The Diagnostic Takeaway**: The 4-profile solution (Model 2) primarily
reflects a unidimensional severity gradient. Because the profiles are
largely parallel across all eight indicators, the differences between
subgroups appear to be quantitative (differences in degree of skills)
rather than qualitative (differences in the pattern of skills). This
suggests that within this 4-class framework, the latent space is defined
by a singular 'general reading ability' factor where students are
distinguished by their overall developmental level.

```{r profile_plots_m2.c4, fig.width=10, fig.height=5, fig.cap="**Figure 5.** *Model 2: 4-Profile Solution (varying variances). Z-Scores relative to sample average (N = 1504)*"}

# ---Get the Model Results ---
mod2_fitted <- scaled_scores_na %>%
  estimate_profiles(1:5, package = "mclust", 
                    variances = "varying", covariances = "zero")

model2_4class <- mod2_fitted[[4]]

# ---Get Estimates and Calculate % ---
plot_data_4 <- get_estimates(model2_4class) %>%
  dplyr::filter(Category == "Means") %>%
  mutate(Parameter = factor(Parameter, levels = order))

# Calculate Percentages for the legend
counts_4 <- table(get_data(model2_4class)$Class)
pcts_4 <- round(prop.table(counts_4) * 100, 1)

# --- Create the Plot ---
ggplot(plot_data_4, aes(x = Parameter, y = Estimate, group = factor(Class), color = factor(Class))) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_errorbar(aes(ymin = Estimate - 1.96*se, ymax = Estimate + 1.96*se), width = 0.1, alpha = 0.3) +
  scale_x_discrete(labels = function(x) str_replace_all(x, "_theta|_score", "")) +
  scale_color_manual(
    values = c("1"="#2ca02c", "2"="#d62728", "3"="#1f77b4", "4"="#ff7f0e"),
    labels = paste0("Class ", names(pcts_4), " (", pcts_4, "%)")
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, face = "bold"),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(y = "Mean Z-Score (relative to average)",
       x = "ROAR Measures")

```

##### Five profiles from Model 2 (varying var)

Figure below displays the five profiles from Model 2 (varying variances,
zero covariances). Unlike the 4-profile solution, the colored lines here
exhibit significant **non-parallelism**, particularly within the
lower-performing subgroups.

**Highlights**: The plots reveal both levels of severity and
**qualitative differences in skill configurations**. In addition to the
proficient (green), above-average (blue), and low-average (purple)
groups, the model identifies two distinct types of struggling readers: a
**specific syntactic-foundational risk** group (orange, 2.9%) and a
**global reading risk** group (Red, 13.4%).

**The Diagnostic Takeaway:** The 5-profile solution reveals a more
complex **structural pattern** than the 4-profile model. While the
higher-performing groups continue to follow a "level-based" trend, the
emergence of the orange profile represents a **qualitative shift** in
the data. This subgroup is characterized by a **discrepant pattern**:
these students demonstrate relatively average performance on the
sentence tasks but show severe, disproportionate "valleys" in receptive
grammar (TROG) and phonological decoding (phonics).

This suggests that reading risk in this population is not purely a
matter of being "behind" in every skill. By moving to a 5-class
framework, we transition from a model of **global developmental delay**
to one that detects a small group of students with **specific needs**.
This granularity is critical for instructional utility, as it
distinguishes students who require a general increase in support from
those who require specialized interventions targeting specific
linguistic and phonological hurdles.

```{r profile_plots_m2.c5, fig.width=10, fig.height=5, fig.cap="**Figure 6.** *Model 2: 5-Profile Solution (varying variances). Z-Scores relative to sample average (N = 1504)*"}

# --- Extract the 5rd model in your mod2_fitted list ---
model2_5class <- mod2_fitted[[5]] 

# --- Prepare the Data ---
plot_data_5 <- get_estimates(model2_5class) %>%
  dplyr::filter(Category == "Means") %>%
  mutate(Parameter = factor(Parameter, levels = order))

# Calculate Percentages for the legend
counts_5 <- table(get_data(model2_5class)$Class)
pcts_5 <- round(prop.table(counts_5) * 100, 1)

# Create the Plot

m2c5.plot <- ggplot(plot_data_5, 
       aes(x = Parameter, y = Estimate, 
           group = factor(Class), color = factor(Class))) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_errorbar(aes(ymin = Estimate - 1.96*se, ymax = Estimate + 1.96*se), 
                width = 0.1, alpha = 0.3) +
  scale_x_discrete(labels = function(x) str_replace_all(x, "_theta|_score", "")) +
  scale_color_manual(
    # FIXED: Removed the trailing comma after 'orchid4'
    values = c("1"="#2ca02c", "2"="#d62728", "3"="#ff7f0e", "4"="#1f77b4", "5"="orchid4"),
    labels = paste0("Class ", names(pcts_5), " (", pcts_5, "%)")
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, face = "bold"),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(y = "Mean Z-Score (Relative to Average)",
       x = "ROAR Measures")

m2c5.plot

```

### Cross-Model Convergence

#### Cross Model Convergence

Figure 7 below presents the VVE 3-profile solution alongside the Model 2
5-profile solution. Notably, a similar configuration of dual weaknesses
emerges in both models, which is characterized by lower performance in
receptive grammar (TROG) and phonological decoding (Phonics) relative to
sentence-level skills.

In the VVE solution, this pattern appears within the lowest-performing
profile (red line), whereas in the 5-profile Model 2 solution it is
isolated as a more distinct subgroup (orange). The recurrence of this
feature across modeling frameworks (mclust vs. tidyLPA) and covariance
specifications (VVE vs. diagonal Model 2) suggests that this contrast
reflects a stable aspect of the observed score structure rather than a
model-specific artifact.

At the same time, differences in class separation and profile
granularity indicate that this pattern is embedded within a broader
severity dimension. Thus, the 5-profile solution can be understood as
providing a more fine-grained partitioning of variation that is already
present, but less differentiated, in the more parsimonious 3-profile
model.

```{r figure-comparison, , fig.width=15, fig.height=7, fig.cap="**Figure 7.** *VVE 3-classe solution vs. Model 2 (VVI) 3-class solution*"}

library(ggplot2)
library(patchwork)
library(stringr)

# 1. Create a shared Y-axis limit so the gridlines match exactly
# Adjust these numbers based on your data's actual min/max
y_limits <- c(-2.5, 1.5) 

# 2. Refine the Model 2 Plot
m2_fixed <- ggplot(plot_data_5, aes(x = Parameter, y = Estimate, group = factor(Class), color = factor(Class))) +
  geom_point(size = 2.5) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", alpha = 0.4) +
  geom_errorbar(aes(ymin = Estimate - 1.96*se, ymax = Estimate + 1.96*se), width = 0.1, alpha = 0.3) +
  # Sync the Y-axis across both plots
  coord_cartesian(ylim = y_limits) +
  scale_x_discrete(labels = function(x) str_replace_all(x, "_theta|_score", "")) +
  scale_color_manual(
    values = c("1"="#2ca02c", "2"="#d62728", "3"="#ff7f0e", "4"="#1f77b4", "5"="#9467bd"),
    labels = paste0("Class ", names(pcts_5), " (", pcts_5, "%)")
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1), 
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank() 
  ) +
  labs(
    title = "Model 2 (VVI): 5-Profile Solution",
    x = "ROAR Measures"
  ) +
  guides(color = guide_legend(ncol = 2, override.aes = list(size = 3)))

# 3. Apply similar limits/cleanup to the first plot (vve.c3.titles)
vve_clean <- vve.c3.titles + 
  coord_cartesian(ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(y = "Mean Z-Score")

# 4. Combine with Patchwork
combined_final <- (vve_clean | m2_fixed) + 
  plot_layout(widths = c(1, 1)) & 
  theme(
    plot.margin = margin(10, 10, 10, 10),
    legend.box.margin = margin(t = -10) # Pulls legend closer to plot
  ) 

combined_final

```

#### Cross-Model Classification Analysis

To evaluate the extent to which the VVE 3-profile solution and the Model
2 5-profile solution recover similar latent groupings, we conducted a
cross-model classification analysis based on individual-level class
assignments. Overall agreement between the two solutions was modest
(Adjusted Rand Index = .32), indicating that the models do not identify
identical partitions of the sample.

Particular attention was given to the apparent similarity between VVE
Class 3 and Model 2 Class 3 with the double weaknesses in the profile
plots. Despite their visual resemblance, cross-model classification
analyses indicated that these classes do not represent the same subgroup
of individuals.

Specifically, Model 2 Class 3 was composed primarily of individuals from
the middle-performing VVE class (72.7%), with a smaller contribution
from the lowest-performing VVE class (27.3%), and no representation from
the highest-performing class. Conversely, only a negligible proportion
of individuals in VVE Class 3 (approximately 2%) were assigned to Model
2 Class 3; instead, the majority were distributed across other Model 2
classes, particularly Class 5 (52.4%).

These results indicate that Model 2 Class 3 does not correspond to a
direct refinement of the lowest-performing VVE group. Rather, it
reflects a recombination of individuals drawn primarily from the
middle-performing class, augmented by a subset of lower-performing
students. The similarity in profile shape across models therefore
appears to arise from a shared contrast in the data—rather than from a
stable subgroup with consistent membership across model specifications.

```{r classfication_overlap}
# Cross-model classification overlap:

# 1. mclust VVE (3 classes)  vs  tidyLPA Model 2 (5 classes)
vve_class <- mod1$classification

# inspect class sizes
vve_sizes <- prop.table(table(vve_class))

# 2. Extract class assignments from tidyLPA mod2.5 results
# model2_5class <- mod2_fitted[[5]]
data_m2_5 <- get_data(model2_5class)

# Look at names to confirm the class-assignment column
# names(data_m2_5)
m2_class <- data_m2_5$Class

# inspect class sizes
m2_sizes <- prop.table(table(m2_class))

# 3. Basic checks
# length(vve_class)
# length(m2_class)


# 4. Cross-tabulation of class assignments
overlap_tab <- table(VVE = vve_class, Model2_5 = m2_class)

# Raw counts as a data frame
overlap_counts_df <- as.data.frame(overlap_tab)


# 5. Row-wise and column-wise proportions
# Row-wise: within each VVE class, where do cases go in Model 2?
overlap_row_prop <- prop.table(overlap_tab, margin = 1)
# round(overlap_row_prop, 3)

# Column-wise: within each Model 2 class, where did cases come from in VVE?
overlap_col_prop <- prop.table(overlap_tab, margin = 2)
# round(overlap_col_prop, 3)

# Data-frame versions for easier inspection/export
overlap_row_df <- as.data.frame(round(overlap_row_prop, 3))
overlap_col_df <- as.data.frame(round(overlap_col_prop, 3))
# overlap_row_df
# overlap_col_df


# 6. Adjusted Rand Index (overall agreement)
ari <- adjustedRandIndex(vve_class, m2_class)

# Interpretation guide:
# < .20   = weak agreement
# .20-.50 = modest
# .50-.80 = moderate to strong
# > .80   = very strong


# 7. Identify the dominant VVE source for each Model 2 class
# For each Model 2 class, find which VVE class contributes the largest proportion
dominant_vve_for_m2 <- apply(overlap_col_prop, 2, function(x) {
  c(
    dominant_VVE_class = which.max(x),
    dominant_prop = max(x)
  )
})

dominant_vve_for_m2 <- as.data.frame(t(dominant_vve_for_m2))
names(dominant_vve_for_m2) <- c("dominant_VVE_class", "dominant_prop")

dominant_vve_for_m2$Model2_5_class <- colnames(overlap_col_prop)

dominant_vve_for_m2 <- dominant_vve_for_m2 %>%
  mutate(
    dominant_VVE_class = as.integer(dominant_VVE_class),
    dominant_prop = round(as.numeric(dominant_prop), 3)
  ) %>%
  select(Model2_5_class, dominant_VVE_class, dominant_prop)


# 8. Identify the dominant Model 2 destination for each VVE class

dominant_m2_for_vve <- apply(overlap_row_prop, 1, function(x) {
  c(
    dominant_M2_class = which.max(x),
    dominant_prop = max(x)
  )
})

dominant_m2_for_vve <- as.data.frame(t(dominant_m2_for_vve))
names(dominant_m2_for_vve) <- c("dominant_M2_class", "dominant_prop")

dominant_m2_for_vve$VVE_class <- rownames(overlap_row_prop)

dominant_m2_for_vve <- dominant_m2_for_vve %>%
  mutate(
    dominant_M2_class = as.integer(dominant_M2_class),
    dominant_prop = round(as.numeric(dominant_prop), 3)
  ) %>%
  select(VVE_class, dominant_M2_class, dominant_prop)


# 9. a compact summary object for reporting

overlap_summary <- list(
  ari = ari,
  counts = overlap_tab,
  row_prop = round(overlap_row_prop, 3),
  col_prop = round(overlap_col_prop, 3),
  dominant_vve_for_m2 = dominant_vve_for_m2,
  dominant_m2_for_vve = dominant_m2_for_vve
)

overlap_summary

# TODO
# 1) Examine profiles separately by grade
# 2) Dimensionality analysis

```

### Multi-group LPA

We conducted the multigroup Latent Profile Analysis (MGLPA) to examine
whether the 5-profile structure is invarianet between two grade-bands:
grades 2-3 and grades 4-5. The MGLPA results are summarized in Table 2.
The 5-profile solution demonstrated robust statistical fit and high
classification clarity (Entropy = 0.83). The Bootstrap Likelihood Ratio
Test ($p < .01$) confirmed that the 5-profile structure significantly
outperformed the 4-profile model. Notably, the smallest identified
subgroup represented 8.7% of the total sample, ensuring sufficient group
stability for comparing developmental patterns across grade-bands.

```{r data_prep_multi}
# Create the two developmental groups
df_groups <- df %>%
  filter(user_grade_at_run < 6) %>% 
  mutate(grade_group = ifelse(user_grade_at_run <= 3, 
                              "Grades 2-3", "Grades 4-5"))

# Standardize WITHIN each group 
# (Critical for developmental comparison)
df_std_grouped <- df_groups %>%
  group_by(grade_group) %>%
  mutate(across(all_of(order), ~as.numeric(scale(.)))) %>%
  ungroup()

dat_numeric <- df_std_grouped %>%
  select(all_of(order))

group_vector <- df_std_grouped$grade_group
```

```{r multiG_LPA_fits}
library(tidyLPA)
# Estimate the 5-profile model across groups
# We use 'estimate_profiles' with the 'group' argument
mg_lpa <- dat_numeric %>%
  estimate_profiles(n_profiles = 5, 
                    package = "mclust", 
                    model_names = "VVI",
                    group = group_vector) 

# Get results
fit_stats <- get_fit(mg_lpa) 


library(dplyr)
library(kableExtra)

# Select and clean the columns for the report
mg_table_clean <- fit_stats %>%
  select(Classes, LogLik, BIC, Entropy, n_min, BLRT_p) %>%
  mutate(
    # Convert n_min to a more readable percentage
    n_min = paste0(round(n_min * 100, 1), "%"),
    # Round the large stats
    across(c(LogLik, BIC, Entropy), ~round(., 2))
  )

# Create the professional HTML table
mg_table_clean %>%
  kbl(caption = "Table 2. Fit Indices for Multigroup Latent Profile Analysis (N=1,504)",
      col.names = c("Profiles", "Log-Likelihood", "BIC", "Entropy", "Smallest Class %", "BLRT (p)"),
      align = "c") %>%
  kable_classic(full_width = F, html_font = "Arial") %>%
  # Adding a note about the multigroup structure
  footnote(general = "Note: Data were standardized within grade-bands (Grades 2-3 vs. 4-5) prior to analysis. 
           BIC: Bayesian Information Criterion; BLRT: Bootstrap Likelihood Ratio Test.")

```

#### Profile Plots

While the 5-profile structure remained statistically stable across
grade-bands, a qualitative comparison of profile shapes revealed a
notable developmental shift in the nature of the most profound reading
difficulty.

```{r mg_profile_line_plots}

mg_data <- get_data(mg_lpa)
mg_data$group <- group_vector

# --- 2. Manually calculate M & se for each Class WITHIN each Group
mg_plot_data_with_se <- mg_data %>%
  # Pivot assessment columns to a long format
  pivot_longer(cols = all_of(order), names_to = "Parameter", values_to = "Score") %>%
  group_by(group, Class, Parameter) %>%
  summarise(
    Mean_Estimate = mean(Score, na.rm = TRUE),
    # Calculate Standard Error: SD / sqrt(n)
    se = sd(Score, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  # Re-apply your expert storytelling order
  mutate(Parameter = factor(Parameter, levels = order))

# --- 3. Plot with Faceting by Grade Group ---
# Define colors to match your bar chart (2 = Red/Struggling)
mg_colors <- c("1"="#1f77b4", "2"="#d62728", "3"="#9467bd", "4"="#ff7f0e", "5"="#2ca02c")

ggplot(mg_plot_data_with_se, aes(x = Parameter, y = Mean_Estimate, 
                                 group = factor(Class), color = factor(Class))) +
  # Add Error Bars FIRST so they sit behind the points
  geom_errorbar(aes(ymin = Mean_Estimate - 1.96*se, 
                    ymax = Mean_Estimate + 1.96*se), 
                width = 0.2, alpha = 0.4) + 
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  facet_wrap(~group) + 
  scale_x_discrete(labels = function(x) stringr::str_replace_all(x, "_theta|_score", "")) +
  scale_color_manual(values = mg_colors) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, face = "bold"),
    legend.position = "none",
    strip.background = element_rect(fill = "gray95"),
    strip.text = element_text(face = "bold", size = 12)
  ) +
  labs(
    title = "Reading Profiles: Developmental Comparison with 95% CI",
    y = "Mean Z-Score (Within-Group)",
    x = NULL
  )
```

In the primary grades (2-3), the lowest performing class (Red Profile)
was characterized by an 'orthographic bottleneck,' where phonological
decoding (Phonics) represented the most profound deficit (Z≈−2.0Z
\approx -2.0Z≈−2.0). In contrast, by upper elementary (4-5), the
weakness of the lowest class shifted toward a 'comprehension
bottleneck.' For these older struggling readers, the most severe
weaknesses were observed in academic vocabulary (CVA) and morphological
knowledge, while decoding skills showed relative maturation.

These patterns appear to align with the common notion that as students
progress through the grades, the primary hurdle for the most at-risk
learners evolves from a failure to master the alphabetic code to a
failure to keep pace with the increasing linguistic and comprehension
demands.

The absence of the specific syntactic-phonological "double-weakness"
profile in the by-grade analysis is likely a function of statistical
power and profile dilution. This highly specific subgroup represented
only 2.9% of the total grades 2-5 sample, and splitting the data into
two grade bands reduced the $N$ for such a rare phenotype below the
threshold for the model to reliably identify it as a distinct class.
Instead, these students were likely re-absorbed into other lower
profiles.

```{r prevalence}
# Create the prevalence cross-tab
prevalence_summary <- mg_data %>%
  group_by(group, Class) %>%
  tally() %>%
  group_by(group) %>%
  mutate(Prevalence_Pct = round(n / sum(n) * 100, 1)) %>%
  arrange(Class, group)

print(prevalence_summary)
```

#### Profile Prevalence

The prevalence of the five classes within each grade-bands indicated a
probable polarization between high- and low-performing students as they
transition from primary to upper-elementary grades.

**Increase at the Top**: The Proficient group (Green) represents a
larger share of the population in Grades 4–5 (35.6%) compared to Grades
2–3 (31.4%), suggesting an upward trajectory.

**Cumulative Risk at the Bottom**: Conversely, the proportion of
students in the two lowest profiles—Below Average and Far-below
Average—increases slightly from a combined 31.7% in the younger group to
37.4% in the older group.

**The Shrinking Middle**: The "Middle" two groups (Above Average and Low
Average) collectively shrink by nearly 10 percentage points.

**The Diagnostic Takeaway**: These results suggest that reading profiles
in our sample may become **increasingly polarized** as students progress
through the grades. It is important to note that because these are
**cross-sectional data**, we cannot definitively infer longitudinal
growth trajectories for individual students from this analysis alone.
**However, the distributional shift is notable**: while a large portion
of the older cohort appears to have shifted toward proficiency, a
substantial and growing subgroup remains 'stuck' or falls further behind
their grade-level peers. This pattern may reflect a widening achievement
gap where early struggles in foundational skills compound over time,
making it increasingly difficult for the lowest-performing students to
keep pace with the rising linguistic demands of the upper-elementary
curriculum.

```{r prevalence_plot}

library(ggplot2)
# 1. Map the colors to the specific Class numbers from your multigroup results

# Order: Class 1 (Green), Class 4 (Blue), Class 5 (Purple), Class 3 (Orange), Class 2 (Red)
mg_colors <- c(
  "1" = "#2ca02c", # Proficient
  "4" = "#1f77b4", # Above Average
  "5" = "#9467bd", # Low Average
  "3" = "#ff7f0e", # Below Average
  "2" = "#d62728"  # Far-below Average
)

library(ggplot2)

# 1. Define the named vector for labels
expert_labels <- c(
  "1" = "Proficient",
  "4" = "Above Average",
  "5" = "Low Average",
  "3" = "Below Average",
  "2" = "Far-below Average"
)

# 2. Define the scholarly 'High-to-Low' order for the X-axis
expert_x_order <- c("1", "4", "5", "3", "2")

# 3. Create the aligned bar chart
ggplot(prevalence_summary, aes(x = factor(Class, levels = expert_x_order), 
                               y = Prevalence_Pct, 
                               fill = factor(Class), 
                               alpha = group)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), 
           width = 0.7, color = "white", show.legend = FALSE) +
  
  # Data labels on top of bars
  geom_text(aes(label = paste0(round(Prevalence_Pct, 1), "%")), 
            position = position_dodge(width = 0.8), 
            vjust = -0.5, size = 3.5, fontface = "bold") +
  
  # Map the COLORS
  scale_fill_manual(values = mg_colors) +
  
  # Map the LABELS to the X-axis
  scale_x_discrete(labels = expert_labels) +
  
  # Map Transparency for Grade Bands
  scale_alpha_manual(values = c("Grades 2-3" = 1.0, "Grades 4-5" = 0.6)) +
  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none", 
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(color = "black"),
    axis.title.x = element_blank() # Remove 'Class' title for a cleaner look
  ) +
  labs(
    title = "Prevalence by Grade Band",
    subtitle = "Solid: Grades 2-3 | Muted: Grades 4-5",
    y = "Percentage of Students (%)"
  ) +
  ylim(0, 45)

```

### Dimensionality Analysis

To evaluate the underlying structure of the eight ROAR measures, we
conducted a Parallel Analysis using the psych package in R. As shown in
the Parallel Analysis Scree Plot, the results indicated a clear
unidimensional structure. The first factor achieved an eigenvalue of
approximately 5.1, significantly exceeding the thresholds generated from
simulated and resampled data. No subsequent factors exceeded the
simulated noise levels.

**The Takeaway**: This statistical evidence suggests that for the
majority of the sample, foundational reading and language comprehension
skills are highly integrated into **a single 'General Reading Ability'
construct**. This unidimensionality accounts for the high degree of
parallelism observed in the primary latent profiles.

However, the emergence of the non-parallel 'Double-Deficit' profile in
higher-order solutions highlights a small but critical subgroup whose
skill configurations deviate from this general developmental trend.

```{r efa}
library(psych)
# 1. Determine how many factors to extract (Parallel Analysis)
# This is the 'BIC plot' equivalent for EFA
fa.parallel(scaled_scores_na, fa = "fa")
```

#### Factor Structure and Construct Integration

To examine the dimensionality of the eight assessment measures, we
conducted an Exploratory Factor Analysis (EFA) with a two-factor target
and oblimin rotation. The results strongly supported a unidimensional
structure. All eight measures loaded heavily onto a single primary
factor (MR1), with loadings ranging from .68 to .91. This single factor
accounted for 64.1% of the total variance in student performance.

```{r factor_structure}

# Run the EFA
# We use 'oblimin' rotation because we expect Language and Code to correlate
efa_results <- fa(scaled_scores_na, 
                  nfactors = 2, rotate = "oblimin")

# 3. Look at the loadings
# print(efa_results$loadings, cutoff = 0)

# Publicationr ready table

library(dplyr)
library(kableExtra)
library(stringr)

# 1. Create the data frame from your output
efa_table_data <- data.frame(
  Measure = c("Word", "Phonics", "Phoneme", "Sentence", 
              "Morphology", "Vocabulary", "TROG", "Inference"),
  MR1 = c(0.815, 0.782, 0.704, 0.772, 0.910, 0.873, 0.680, 0.840),
  MR2 = c(-0.153, 0.200, 0.207, -0.292, -0.012, -0.022, 0.249, -0.006)
)

# 2. Add the Variance/SS Loading rows at the bottom
summary_rows <- data.frame(
  Measure = c("SS Loadings (Eigenvalues)", "Proportion Variance", "Cumulative Variance"),
  MR1 = c(5.125, 0.641, 0.641),
  MR2 = c(0.254, 0.032, 0.672)
)

final_table <- bind_rows(efa_table_data, summary_rows)

# 3. Create the HTML Table
final_table %>%
  kbl(caption = "Table 2. Exploratory Factor Analysis Loadings and Variance Explained",
      col.names = c("Assessment Measure", "Factor 1 (General Reading)", "Factor 2"),
      digits = 3,
      align = "lcc") %>%
  kable_classic(full_width = F, html_font = "Arial") %>%
  # Bold the significant loadings in Factor 1
  column_spec(2, bold = ifelse(final_table$MR1 > 0.4, TRUE, FALSE)) %>%
  # Add a line to separate measures from statistics
  row_spec(8, extra_css = "border-bottom: 1px solid;") %>%
  footnote(general = "Extraction Method: Minimum Residual. Rotation: Oblimin. Loadings > .40 are bolded. Factor 1 represents a unified General Reading Ability construct.")

```

A second factor (MR2) failed to achieve statistical or substantive
significance, explaining only 3.2% of additional variance with no
variables loading above the .30 threshold.

While the earlier Parallel Analysis and Eigenvalues ($SS < 1.0$) confirm
a unidimensional structure, the residual loadings in the non-significant
second factor (MR2) provide an interesting diagnostic information we
observed with the latent profile analysis with the full grades 2-5
sample.

Specifically, the variables that showed the most variance in the latent
profile analysis—receptive grammar (TROG) and phonological
decoding—showed a shared directional pull in MR2 ($loadings \approx .20$
to $.25$) that contrasted with sentence-level performance
($loading \approx -.29$).

This suggests that although reading ability is 64% unidimensional in
this sample, the primary axis of 'divergence' for struggling readers
involves a split between structural linguistic/phonological skills and
contextual sentence processing.

This provides the structural justification for our use of Model 2
(varying variances) in the LPA, as it allows the model to capture this
specific pattern of divergence that the general factor otherwise
obscures.

### Within-person Centering

**The Logic**: Currently, most profiles are "parallel" because the
general reading ability accounts for 64% of the variance (as seen in
the EFA results). By subtracting each student's personal average from their own
scores, qw force every student to have a "personal mean of zero."

**How it works**: A student who is high in everything looks the same as
a student who is low in everything (both are flat lines at 0).

**Benefit**: Only students with discrepancies (e.g., high phonics/low
grammar) will show peaks and valleys. Effectively we are zooming on the zig zag lines (valleys and peaks within students).

```{r ipsative_data_prep}

# 1. Start with your standardized data
# 2. Calculate a "Personal Mean" for each row and subtract it
ipsative_data <- scaled_scores_na %>%
  rowwise() %>%
  mutate(personal_mean = mean(c_across(everything()))) %>%
  mutate(across(everything() & -personal_mean, ~ . - personal_mean)) %>%
  select(-personal_mean) %>%
  ungroup()

```


```{r tidyLPA_mod_com_ipsative}

library(tidyLPA)
library(dplyr)
library(kableExtra)

# Note: In mclust terms: 
# Model 1 = EEI (Equal variance, diagonal)
# Model 2 = VVI (Varying variance, diagonal)

# --- Run Model 1 (Equal Variance) ---
mod1_results <- ipsative_data %>%
  estimate_profiles(1:5, package = "mclust", 
                    variances = "equal", 
                    covariances = "zero") %>%
  get_fit() %>%
  mutate(Model_Type = "Model 1 (Equal Var)")

# --- Run Model 2 (Varying Variance) ---
mod2_results <- ipsative_data %>%
  estimate_profiles(1:5, package = "mclust", 
                    variances = "varying", 
                    covariances = "zero") %>%
  get_fit() %>%
  mutate(Model_Type = "Model 2 (Varying Var)")


# --- Combine into one Master Comparison Table ---
all_comparisons <- bind_rows(mod1_results, mod2_results)

all_comparisons_short <- all_comparisons %>%
  select(Model_Type, Classes, n_parm = parameters,
         BIC, Entropy, n_min, BLRT_p)

# all_comparisons_short <- bind_rows(
#   all_comparisons_short,
#   vve_fit_row
# )

# --- Clean and Present ---
all_comparisons_short %>%
  mutate(across(c(BIC, Entropy, n_min), ~round(., 2))) %>%
  arrange(Model_Type, Classes) %>% 
  kbl(caption = "Table 1a. Sensitivity Analysis: Comparing Profile Solutions across Model Specifications (Z-Scores) with the Ipsative Data") %>%
  kable_classic(full_width = F, html_font = "Arial") 
# %>%
#   # Highlight all 3-profile solutions to compare the 'Winner' across specifications
#   row_spec(which(all_comparisons$Classes == 3), bold = T, background = "#f9f9f9")

```
<br><br>
##### Five profiles from Model 2 (varying var) with Within-person Centered Data

This plot displays mean relative scores (within-person centering), which removes global achievement levels to focus on the configural shape (relative strengths and weaknesses) of each student subgroup.

Figure 7 shows that the 'Balanced Anchor' group (Class 3; 29.3%) exhibited no relative discrepancies with a relatively flat orange line. In contrast, the remaining 70% of the sample demonstrated notable zig-zag lines, indicating uneven component skill development. 

Notably, the Green (Class 1) and Purple (Class 5) profiles revealed a distinct pattern of structural weaknesses.

These students demonstrated a relative peak in sentence efficiency, indicating a preserved ability to quickly process and verify a simple sentence (e.g., "The sky is blue"). However, they exhibited disproportionate "valleys" in receptive grammar (TROG) and phonological decoding (Phonics). In the case of the Purple profile (5.7%), this gap was extreme, signifying a clear need for dual-focus support.

This specific configuration suggests that while these learners have achieved a degree of functional automaticity for simple, familiar text, they lack the generative tools—specifically syntactic processing and phonological assembly—required to handle more complex or unfamiliar literacy tasks. 

This finding highlights a critical instructional priority: students who appear efficient on basic sentence-level fluency measures may nonetheless possess significant underlying structural weaknesses that require integrated support across both linguistic and orthographic domains.

```{r profile_plots_m2.c5_ipsative_data, fig.width=8, fig.height=5, fig.cap="**Figure 7. **Five Reading Mean Profile Plot with Within-person Centered Data (N = 1504)**"}

# --- Run Model 2 5 Profiles
# ---Get the Model Results ---
mod2_fitted <- ipsative_data %>%
  estimate_profiles(1:5, package = "mclust", 
                    variances = "varying", covariances = "zero")

# --- Extract the 5rd model in your mod2_fitted list ---
model2_5class <- mod2_fitted[[5]] 

# --- Prepare the Data ---
plot_data_5 <- get_estimates(model2_5class) %>%
  dplyr::filter(Category == "Means") %>%
  mutate(Parameter = factor(Parameter, levels = order))

# Calculate Percentages for the legend
counts_5 <- table(get_data(model2_5class)$Class)
pcts_5 <- round(prop.table(counts_5) * 100, 1)

# Create the Plot

m2c5.plot <- ggplot(plot_data_5, 
       aes(x = Parameter, y = Estimate, 
           group = factor(Class), color = factor(Class))) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_errorbar(aes(ymin = Estimate - 1.96*se, ymax = Estimate + 1.96*se), 
                width = 0.1, alpha = 0.3) +
  scale_x_discrete(labels = function(x) str_replace_all(x, "_theta|_score", "")) +
  scale_color_manual(
    # FIXED: Removed the trailing comma after 'orchid4'
    values = c("1"="#2ca02c", "2"="#d62728", "3"="#ff7f0e", "4"="#1f77b4", "5"="orchid4"),
    labels = paste0("Class ", names(pcts_5), " (", pcts_5, "%)")
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, face = "bold"),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(y = "Mean Z-Score (Relative to each student's mean)",
       x = "ROAR Measures",
       subtitle = "using mean relative scores (de-meaned Z-scores)")

m2c5.plot

```

### Filter by Intra-Individual Variance

**The Logic**: Many students are "consistent"—their scores across the 8 tests are very similar. These students create those parallel "Level" profiles. So we first identify these "Consistent" students, set them aside, and then run the LPA only on the "Inconsistent" students.

**How it works**: If a student's scores are all very close to each other, their variance will be low (variance < .25). If they have jagged scores, their variance will be high. We will analyze a subset of students who have score variance higher than 0.25. 

Mathematically, an intra-individual variance of 0.25 corresponds to an average deviation of 0.50 standard deviations from a student’s own mean across the eight assessments. In the context of eight ROAR-measures, this 'scatter' may shed light on a substantive decoupling of foundation component and linguistic comprehension skills. 

```{r filter_consistent_students}
# 1. Calculate variance for each student across the 8 scores
df_with_var <- scaled_scores_na %>%
  rowwise() %>%
  mutate(indiv_var = var(c_across(everything()))) %>%
  ungroup()

# library(summarytools)
view(dfSummary(df_with_var))


# 2. Split the sample
consistent_students <- df_with_var %>% filter(indiv_var < 0.25)
jagged_students <- df_with_var %>% filter(indiv_var >= 0.25)

# 3. Report the 'Consistent' group descriptively (e.g., "70% of students show flat profiles")
# Calculate stats for the Consistent group
n_total <- nrow(df_with_var)
n_consistent <- nrow(consistent_students)
pct_consistent <- round((n_consistent / n_total) * 100, 1)

# --- 1. Consistent Profiles Summary ---
n_consistent <- nrow(consistent_students)
pct_consistent <- round((n_consistent / n_total) * 100, 1)
avg_level_consistent <- mean(rowMeans(consistent_students %>% select(all_of(order))))

cat("Descriptive Summary of Consistent Profiles:\n",
    "------------------------------------------\n",
    "Total Sample N:", n_total, "\n",
    "Consistent Students (Var < 0.25):", n_consistent, "(", pct_consistent, "% of sample)\n",
    "Average Z-score (Global Level):", round(avg_level_consistent, 2), "\n\n")

# --- 2. Jagged Profiles Summary ---
# Calculate the specific N and % for the jagged group
n_jagged <- nrow(jagged_students)
pct_jagged <- round((n_jagged / n_total) * 100, 1)

# Calculate the mean 'Global Ability' for the jagged group
avg_level_jagged <- mean(rowMeans(jagged_students %>% select(all_of(order))))

cat("Descriptive Summary of Jagged Profiles:\n",
    "------------------------------------------\n",
    "Total Sample N:", n_total, "\n",
    "Jagged Students (Var >= 0.25):", n_jagged, "(", pct_jagged, "% of sample)\n",
    "Average Z-score (Global Level):", round(avg_level_jagged, 2), "\n")

```

<br><br>
#### Five profiles from Model 2 (varying var) with a Subset of Students with a Large Variance with the 8 ROAR scores

While the within-person centered analysis (Figure 7) isolated relative skill configurations across the total sample, we conducted a follow-up LPA on the high-variance subsample ($n = 837$) to examine how these configurations manifest when absolute achievement levels are preserved (Figure 8).

In Figure 8, we observe that for students in the proficient and average ranges (Classes 5, 4 and 2), skill configurations remain relatively stable despite meeting the threshold for high intra-individual variance. However, at the lower end of the achievement spectrum, the profiles become markedly non-parallel. 

Specifically, the Dual-Focus Support group (Orange, Class 3, 8.7%) emerges as a distinct 'valley-shaped' profile, where receptive grammar and phonics gaps are disproportionately severe relative to their other skills. This demonstrates that non-parallel skill development is a hallmark of low performance in this sample: as the overall reading ability decreases, the imbalanced skill mastery pattern emerges. These students demonstrated relative preservation of sentence-level processing alongside profound, specific instructional needs in receptive grammar (Z ≈ -2.0) and phonological decoding (Z ≈ -1.8). 

By isolating the 'jagged/inconsistent' students from those with balanced profiles, we demonstrate that the syntactic-phonological gap is not merely a rare exception, but a prominent pattern for students who may be struggling to co-develop linguistic and foundational skills for proficient reading.

```{r profile_plots_m2.c5_jaggered, fig.width=8, fig.height=5, fig.cap="**Figure 8. **Reading Profiles Among Students with High Intra-Individual Skill Variance (Model 2, n = 837)**"}


##### Five profiles from Model 2 (varying var) with Ipsative Data

# --- Run Model 2 5 Profiles
# ---Get the Model Results ---
mod2_fitted <- jagged_students %>%
  estimate_profiles(1:5, package = "mclust", 
                    variances = "varying", covariances = "zero")

# --- Extract the 5rd model in your mod2_fitted list ---
model2_5class <- mod2_fitted[[5]] 

# --- Prepare the Data ---
plot_data_5_clean <- get_estimates(model2_5class) %>%
  dplyr::filter(Category == "Means") %>%
  # NEW: Keep only rows where Parameter is one of your 8 assessments
  dplyr::filter(Parameter %in% order) %>% 
  # NEW: Explicitly drop any "ghost" factor levels
  mutate(Parameter = factor(Parameter, levels = order)) %>%
  droplevels() 

# Check the result
# unique(plot_data_5_clean$Parameter) # no NA

# Calculate Percentages for the legend
class_counts <- table(get_data(model2_5class)$Class)
pcts_5 <- round(prop.table(class_counts) * 100, 1)

# Create the Plot

m2c5.plot <- ggplot(plot_data_5_clean, 
       aes(x = Parameter, y = Estimate, 
           group = factor(Class), color = factor(Class))) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_errorbar(aes(ymin = Estimate - 1.96*se, ymax = Estimate + 1.96*se), 
                width = 0.1, alpha = 0.3) +
  scale_x_discrete(labels = function(x) str_replace_all(x, "_theta|_score", "")) +
  scale_color_manual(
    # FIXED: Removed the trailing comma after 'orchid4'
    values = c("1"="#2ca02c", "2"="#d62728", "3"="#ff7f0e", "4"="#1f77b4", "5"="orchid4"),
    labels = paste0("Class ", names(pcts_5), " (", pcts_5, "%)")
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, face = "bold"),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(y = "Mean Z-Score (relative to sample mean)",
       x = "ROAR Measures",
       subtitle = "using subset of learners exhibiting significant skill imbalance (variance > 0.25)")

m2c5.plot

```

#### Note on Different Y-range in Figures 7 and 8

The scales were purposefully allowed to vary to reflect the different nature of the data being visualized. Figure 7 employs a **zoom-in lens** approach to highlight configural shapes; using a wider normative scale would compress these qualitative trade-offs and render the diagnostic 'valleys' invisible. 

Conversely, Figure 8 uses a **wide-angle lens** to illustrate how these qualitative shapes exist within the broader context of absolute achievement levels while focusing on those students with a large variance, indicating inconsistent levels of component skill mastery.

Maintaining separate scales is essential to accurately represent the discrepancy between relative instructional needs and absolute performance goals.

### Problem with Standardized Scores

```{r plots1}
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

# --- 1. Simulation ---
set.seed(42)
n <- 1000

# The Master Ruler (DDA Scale)
# Phonics: Low mean, very narrow spread
# CVA: High mean, very wide spread
phonics_dda <- rnorm(n, mean = -2.25, sd = 0.4) 
cva_dda     <- rnorm(n, mean = 2.0,   sd = 1.8)

df_dda <- data.frame(Phonics = phonics_dda, CVA = cva_dda)

# 2. Derive Local Scales (Centered at 0, keeping natural SDs)
df_local <- df_dda %>%
  mutate(Phonics = Phonics - mean(Phonics),
         CVA = CVA - mean(CVA))

# 3. Derive Z-Scales (Centered at 0, SD forced to 1)
df_z <- df_dda %>%
  mutate(across(everything(), ~as.numeric(scale(.))))

# --- Plotting ---
p_setup <- function(data, title, xlab) {
  ggplot(data %>% pivot_longer(everything()), aes(x = value, fill = name)) +
    geom_density(alpha = 0.4, color = "white") +
    scale_fill_manual(values = c("Phonics" = "#d62728", "CVA" = "#1f77b4")) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none", plot.title = element_text(face="bold")) +
    labs(title = title, x = xlab, y = "Density")
}

p1 <- p_setup(df_local, "1. Local IRT Scales", "Ability relative to Task Mean (Logits)")
p2 <- p_setup(df_z, "2. Standardized World (Z-Scores)", "Relative Rank (SD Units)")
p3 <- p_setup(df_dda, "3. Aligned World (DDA)", "Shared Master Ruler (Logits)")

(p1 / p2 / p3) + plot_layout(guides = "collect") & theme(legend.position = "bottom", legend.title = element_blank())
```

```{r plots2}
# Creating a "Translation Plot" for Alpha and Delta
trans_df <- data.frame(local_theta = seq(-3, 3, 0.1)) %>%
  mutate(
    # Aligned = (Alpha * Local) + Delta
    # Phonics: Narrow (alpha 0.4), Hard/Low (delta -2.25)
    Phonics_Aligned = (0.4 * local_theta) - 2.25,
    # CVA: Wide (alpha 1.8), Easy/High (delta 2.0)
    CVA_Aligned = (1.8 * local_theta) + 2.0
  ) %>%
  pivot_longer(-local_theta, names_to = "Task", values_to = "Master_Theta")

ggplot(trans_df, aes(x = local_theta, y = Master_Theta, color = Task)) +
  geom_line(linewidth = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) + # The Z-score assumption
  scale_color_manual(values = c("CVA_Aligned" = "#1f77b4", "Phonics_Aligned" = "#d62728")) +
  annotate("text", x = 1, y = 4.5, label = "CVA: Alpha = 1.8 (STRETCH)\nDelta = +2.0 (EASY)", color = "#1f77b4", fontface="bold") +
  annotate("text", x = 1, y = -1.5, label = "Phonics: Alpha = 0.4 (SQUASH)\nDelta = -2.25 (HARD)", color = "#d62728", fontface="bold") +
  theme_minimal() +
  labs(title = "How DDA Rewrites the Ruler",
       subtitle = "Dashed line = Z-score assumption (all tests are equal). \nColored lines = Real developmental exchange rates.",
       x = "Score on the Local Test", y = "Value on the Master Reading Ruler")
```

```{r prob_z_scores_illustration}
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

# --- 1. Simulation Parameters (From your ROAR table) ---
set.seed(123)
n <- 2000
sd_phon <- 1.5  # Phonics SD from your table
sd_cva  <- 2.0  # CVA SD from your table

# --- 2. Defining the "Truth" ---
# Student A: Has a 0.5 logit gap in the local scale.
# Student B: Perfectly balanced in the local scale.
students <- data.frame(
  Name = c("Student A", "Student B"),
  Ref_Phon = c(-1.5, 1.5), 
  Ref_CVA  = c(-2.0, 1.5), # B is balanced at 1.5; A has a 0.5 difference
  y_pos = c(0.05, 0.15)    # Staggered height
) %>%
  mutate(
    # Panel 2: Standardized (Z-Scores)
    # Student A: -1.5/1.5 = -1.0 | -2.0/2.0 = -1.0 (Gap deleted!)
    # Student B: 1.5/1.5 = 1.0 | 1.5/2.0 = 0.75 (Fake gap created!)
    Z_Phon = Ref_Phon / sd_phon, 
    Z_CVA  = Ref_CVA / sd_cva,
    
    # Panel 3: DDA Aligned (Master Ruler)
    # We apply the DDA logic: shift means based on absolute difficulty
    # Phonics is foundational (Mean -1.5), CVA is advanced (Mean +1.5)
    DDA_Phon = Ref_Phon - 1.5, 
    DDA_CVA  = Ref_CVA + 1.5
  )

# --- 3. Plotting Helper ---
clean_theme <- theme_minimal(base_size = 12) + 
  theme(legend.title = element_blank(), # Removes 'Name' and 'name'
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 14))

# PANEL 1: REFERENCE WORLD (Standard IRT)
p1 <- ggplot() +
  geom_density(data = data.frame(x = rnorm(n, 0, sd_phon)), aes(x, fill = "Phonics"), alpha = 0.3, color = NA) +
  geom_density(data = data.frame(x = rnorm(n, 0, sd_cva)), aes(x, fill = "CVA"), alpha = 0.3, color = NA) +
  geom_segment(data = students, aes(x = Ref_Phon, xend = Ref_CVA, y = y_pos, yend = y_pos, color = Name), linewidth = 1.5) +
  geom_point(data = students %>% pivot_longer(cols = c(Ref_Phon, Ref_CVA)), aes(x = value, y = y_pos, color = Name), size = 4) +
  scale_fill_manual(values = c("Phonics" = "#d62728", "CVA" = "#1f77b4")) +
  scale_color_manual(values = c("Student A" = "orange", "Student B" = "purple")) +
  labs(title = "1. Reference World (Local IRT)", 
       subtitle = "Rulers differ in width (SD). Student B is balanced; Student A has a small gap.",
       x = "Ability relative to Task Mean (Logits)", y = "Density") + clean_theme

# PANEL 2: STANDARDIZED WORLD (The Statistical Distortion)
p2 <- ggplot() +
  geom_density(data = data.frame(x = rnorm(n, 0, 1)), aes(x, fill = "Phonics"), alpha = 0.3, color = NA) +
  geom_density(data = data.frame(x = rnorm(n, 0, 1)), aes(x, fill = "CVA"), alpha = 0.3, color = NA) +
  geom_segment(data = students, aes(x = Z_Phon, xend = Z_CVA, y = y_pos, yend = y_pos, color = Name), linewidth = 1.5) +
  geom_point(data = students %>% pivot_longer(cols = c(Z_Phon, Z_CVA)), aes(x = value, y = y_pos, color = Name), size = 4) +
  scale_fill_manual(values = c("Phonics" = "#d62728", "CVA" = "#1f77b4")) +
  scale_color_manual(values = c("Student A" = "orange", "Student B" = "purple")) +
  labs(title = "2. Standardized World (Z-Scores)", 
       subtitle = "Math forces SD=1. Student A's gap is DELETED. Student B's gap is MANUFACTURED.",
       x = "Relative Standing (SD Units)", y = "Density") + clean_theme

# PANEL 3: ALIGNED WORLD (The DDA Solution)
p3 <- ggplot() +
  geom_density(data = data.frame(x = rnorm(n, -1.5, sd_phon)), aes(x, fill = "Phonics"), alpha = 0.3, color = NA) +
  geom_density(data = data.frame(x = rnorm(n, 1.5, sd_cva)), aes(x, fill = "CVA"), alpha = 0.3, color = NA) +
  geom_segment(data = students, aes(x = DDA_Phon, xend = DDA_CVA, y = y_pos, yend = y_pos, color = Name), linewidth = 1.5) +
  geom_point(data = students %>% pivot_longer(cols = c(DDA_Phon, DDA_CVA)), aes(x = value, y = y_pos, color = Name), size = 4) +
  scale_fill_manual(values = c("Phonics" = "#d62728", "CVA" = "#1f77b4")) +
  scale_color_manual(values = c("Student A" = "orange", "Student B" = "purple")) +
  labs(title = "3. Aligned World (Master Ruler via DDA)", 
       subtitle = "Reflects true difficulty distance. Student A's gap is revealed; Student B remains proficient.",
       x = "Shared Growth Scale (Logits)", y = "Density") + clean_theme

# Final Layout
(p1 / p2 / p3) + plot_layout(guides = "collect") & theme(legend.position = "bottom")

```